---
title: "Introduction to R"
author: "ASHIR MEHTA"
date: "9/11/2020"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A brief history of R ##

- S created by John Chambers at Bell Laboratories in 1976
- "Turn ideas into software, quickly and faithfully"
- R created in 1993 as a free, open-source implementation of S
- Influenced by S, Scheme, Common Lisp, XLispStat

## Basic data types (atomic) ##

There are six 'atomic' data types:

- **character**
- **double**
- **integer**
- **logical**
- raw
- complex

Most of the time, R will not distinguish between integer and double, as both are considered *numeric* vectors by R.

----

```{r}
"a"
1
1L
TRUE
```

----

- Everything is a vector in R

- "Scalars" are actually length-1 vectors

- Vectors of atomic types are created using __c()__

- Sequences of integers can be shortcut via __start:end__

- Assignment is done via __<-__

----

```{r}
c("a", "b", "c", "d")
c(1, 2.2, 3.33, 4.444)
c(1L, 2L, 3L, 4L)
c(TRUE, FALSE, TRUE, NA)
```

----

```{r}
x <- c(1, 2, 3, 4)
x
y <- 1:4
y
```

## Basic data types (non-atomic)

Atomic vector types are used to build more complex data types:

- matrix
- array
- list
- data.frame

This class will primarily discuss and make use of data.frames, which are the most common way of storing data in R.

For most types of data, systematic and tidy use of data.frames makes data analysis quick and easy.

----

```{r}
matrix(c(1, 2, 3, 4), nrow=2, ncol=2)
```

----

```{r}
array(1:12, dim=c(2,3,2))
```

----

```{r}
list(1L, "a", 3.14)
```

----

```{r}
data.frame(x=c(1, 2, 3), y=c("a", "b", "c"))
```

----

The table below visualizes the relationship between these basic data types:

R data types | Same types | Different types
-------------|------------|----------------
   1d        |  vector    |  list
   2d        |  matrix    |  data.frame
   nd        |  array     |  <NA>

----

Each of these data types can be subsetted in multiple ways:

R subscripts         | Simplifying    | Preserving
---------------------|----------------|-----------
vectors              | x[[1]]         | x[1:6]
matrices/data.frames | x[,1]          | x[,1,drop=FALSE]
lists                | x[[1]], x$name | x[1]

Note that indexing in R begins with 1, _not_ 0.

Also note that data.frames are actually a special type of list, and can also be subset in the same ways that a list can.

## Subsetting: "simplying" versus "preserving"

When subsetting non-atomic data types (e.g., lists), subsetting operations may be distinguished by whether they return:

- An object of the same type (e.g., another list)

- An simpler object that was contained in the original object (e.g., a vector inside a list)

For example, if `z` is a list:

- `z[1]` returns a list of length 1, including the first element of `z`

- `z[[1]]` returns the first element of `z` (exactly)

----

```{r}
z <- list(1L, "a", 3.14)
z[1:2]
z[1]
z[[1]]
```

----

```{r}
df <- data.frame(x=c(1, 2, 3), y=c("a", "b", "c"))
df$x
df[,"y"]
df[1:2,]
```

## Documentation

- Access help page for any R function using __help(name)__

- Syntactic sugar is available as __?name__

- You may need to quote special names with backticks

## Functions

Functions are created with the `function` keyword, and then assigned to a variable.

```{r}
add <- function(a, b) a + b
add(1, 2)
add(a=1, b=2)
```

----

Functions are first-class, and can be treated as any other variable.

```{r}
add2 <- add
add2(2, 3)
```

----

Functions do not need to be assigned to a variable to be called.

```{r}
(function(a, b) a + b)(1, 2)
```

## Control flow

Control flow in R includes `if`/`else` statements, `while` loops, and `for` loops, and the `switch` function.

```{r}
if (TRUE) {
  print(1)
} else {
  print(2)
}
```

```{r}
i <- 1
while(i < 3) {
  print(i)
  i <- i + 1
}
```


```{r}
for ( i in 1:3 ) {
  print(i)
}

for ( xi in list(1, "x", "y", TRUE) ) {
  print(xi)
}
```

```{r}
x <- "b"
switch(x, a=1, b=2)
y <- 2
switch(y, "a", "b")
```

## Apply functions

Rather than using `for` loops, it is common to use the `*apply` family of functions. These allow applying a function over each element of a vector.

- `lapply` always returns its results as a list.

- `sapply` is a variant of `lapply` that attempts to simplify its final result

- `vapply` is a variant of `lapply` that simplifies its result according to a template.

#### lapply

`lapply` always returns its results as a list.

```{r}
x <- list(1:3, 4:6, 7:9)
lapply(x, sum)
```

#### sapply

`sapply` is a variant of `lapply` that attempts to simplify its final result into a homogenous vector, matrix, or array.

```{r}
x <- list(1:3, 4:6, 7:9)
sapply(x, sum)
```

#### vapply

`vapply` is a variant of `lapply` that simplifies its result according to a user-supplied template.

```{r}
x <- list(1:3, 4:6, 7:9)
vapply(x, sum, numeric(1))
```

## Variable arguments via dot-dot-dot

Many functions in R can take a variable number of arguments, and possibly pass these arguments along to other functions,

This is done using `...` in the function signature.

```{r}
print2 <- function(...) {
  args <- list(...)
  for ( a in args )
    print(a)
}
```

---

```{r}
print2(1, "a", TRUE)
```

## Plotting using "base" graphics

- Base R plotting uses "brush on canvas" philosophy

- Plot elements are plotted individually, one on top of another

- Allows "formula" interface

- Good enough for basic plots

- Simple and flexible

- We will learn a more powerful alternative (ggplot2) later

----

```{r}
x <- seq(from=0, to=2*pi, length.out=100)
sinx <- sin(x)
plot(x, sinx)
```

----

```{r}
df <- data.frame(x=x, y=sinx)
plot(y ~ x, data=df)
```

---

```{r}
x <- rnorm(100)
hist(x)
```

